A Type System for Structured Aliasing in Parallel Programs

### 2012-10-29 doc/types-pldi2013/intro.tex
Adding some references

To combat this problem, researchers have previously explored programming
systems that are capable of reasoning about the structure of data in
parallel programs.  Traditionally, these attempts have taken one of two approaches.   
The first approach relies on static analyses that
have no runtime overhead, but limited expressivity.  Two recent examples, 
Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap recursively into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related} for more discussion of related work).  The
second approach relies on a dynamic analysis of program data that doesn't
limit expressivity, but incurs runtime overhead.  Examples of this approach 
include transactional memory and thread-level speculation.

These two different approaches illustrate different mechanisms for dealing with
the problem of data aliasing in parallel programs.  The fully static approach disallows aliasing
by requiring a single static partitioning of data.  Our own experience with writing high-performance 
applications in both Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data.  Alternatively, 
the fully dynamic approach enables arbitrary data aliasing, but requires 
centralized control logic to perform conflict detection between every pair 
of parallel computations. This results in runtime overhead and is 
not scalable on distributed-memory machines.

>>>>

To combat this problem, researchers have previously explored programming
systems that are capable of reasoning about the structure of data in
parallel programs.  Traditionally, these attempts have taken one of two approaches.   
The first approach relies on fully static analyses that
have no runtime overhead, but limited expressivity.  Two recent examples, 
Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap recursively into a tree of collections 
of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be fully checked statically 
(see Section~\ref{sec:related} for more discussion of related work).  The
second approach relies on a totally dynamic analysis of program data that doesn't
limit expressivity, but incurs runtime overhead.  Examples of this approach 
include transactional memory \cite{Harris05} and thread-level speculation \cite{Steffan00}.

These two approaches illustrate different mechanisms for dealing with
the problem of data aliasing in parallel programs.  The fully static approach disallows aliasing
by requiring a single static partitioning of data.  Our own experience with writing high-performance 
applications in both Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data.  Alternatively, 
the fully dynamic approach enables arbitrary data aliasing, but requires 
centralized control logic to perform conflict detection between every pair 
of parallel computations. This results in runtime overhead and is 
not scalable on distributed-memory machines.
